---
title: "Discarded"
author: "Rakeen Tanvir"
date: "12/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Completed Tasks:

Install GitHub and GitHub Desktop App

Share repo on GitHub to usernames: graydh, arslansaid

# To do

Sampling - empirical data vs. theoretical distribution using chi-square goodness of fit tests, bootstrapping
Monte Carlo - homework #2
  Identify distribution for data
  Investigate whether t-test is appropriate with power study
  Determine confidence interval for variable of interest
Bootstrapping - quantify uncertainty with confidence interval of point estimate (double numeric predictor variables)
Difference in differences?
Interaction/effects of economic variables?

Regression - robust, logistic, interaction
https://www.datacamp.com/community/tutorials/logistic-regression-R
https://daviddalpiaz.github.io/r4sl/logistic-regression.html


At which annual rate of inflation does an Inflation Crisis become a practical certainty?

# Ideas
Simpson's Paradox?
Changes in regression relationships
What variables should be thrown away?
Repeat analysis for variable is T or F, or in or out of model.
Plots + confidence intervals
How are the different crises related? Heat map, refine question. Reverse engineering to identify motivation for compiling the data set. Investigate relevant research and economic definitions. 
Because we are working with time series data, let's maintain continuity over time and split the data by country.

```{r}
plt <- ggplot(data = df)
plt + geom_bar(mapping = aes(x = case))
plt + geom_bar(mapping = aes(x = cc3))
plt + geom_bar(mapping = aes(x = country)) +
  theme(axis.text.x=element_blank())

(factors <- unique(df[ , 1:3]) )
```

### Split the data into three pieces

~60% of the data goes into a training (or exploration) set

~20% goes into a test set used to compare models by hand

~20% is held back for a validation set to test the final model

What split decision rule should we follow?

1. Split by country.

2. Randomly select observations from the four highlighted countries to split among the query and test sets. 

3. Randomly select observations from the entire data set to split among the training, query, and test sets. 

## Preliminary checks on data quality

### Edit checks and plots to identify gross data errors and extreme outliers

### Diagnostics for relationships and strong interactions

### Apply remedial measures, if needed


# Reduction of number of explanatory variables (for exploratory observational studies)

Not required for confirmatory observational studies. 

## Determine several potentially useful subsets of explanatory variables; include known essential variables


# Model refinement and selection

## Investigate curvature and interaction effects more fully

## Study residuals and other diagnostics

## Apply remedial measures, if needed

## Select tentative model


# Model validation - 2-3 data sets

## Validity checks

## Final regression model

### View the factor variables with plots and a tibble