---
title: 'STAT 535 Final Project - Fall 2021'
date: 'Due Date: December 15, 2021, 5:59pm'
author: "Names: Said Arslan, Declan Gray-Mullen, and Rakeen Tanvir"
output: 
  html_document:
    toc: TRUE
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
```

Completed Tasks:
Install GitHub and GitHub Desktop App
Share repo on GitHub to usernames: graydh, arslansaid

Abstract:
Predictive Model of Systemic Crises in Africa

• Full names of group members
Said Arslan
Declan Gray-Mullen
Rakeen Tanvir

• A brief description of the project and data collection (website, date type etc.)
https://www.kaggle.com/chirin/africa-economic-banking-and-systemic-crisis-data 


• Objectives and interests (what will be predicted)
How debt affects crises? 
Interaction/effects of economic variables?
Difference in differences?
Regression - robust, logistic, interaction
At which annual rate of inflation does an Inflation Crisis become a practical certainty?

Deeper dive into variables and data sources/definitions
  Data collection
  Data manipulation

• Class material to be used for the analysis
RMarkdown
Data Types - lists, data frames, tibbles 
Flow Control - if, else, while, repeat
Strings & Regular Expression - character (country code, country)
  Web scraping - aggregate data on African countries
Data Transformations (dplyr & SQL)
Visualization (ggplot) - regression, inflation CPI vs. year, country, etc. 
Factors - country
Model fitting/optimization - regression
Functions of functions - helper functions, custom functions for regression
Sampling - empirical data vs. theoretical distribution using chi-square goodness of fit tests, bootstrapping
Monte Carlo - homework #2
  Identify distribution for data
  Investigate whether t-test is appropriate with power study
  Determine confidence interval for variable of interest
Bootstrapping - quantify uncertainty with confidence interval of point estimate

• A description of the responsibilities of each of the group members.
Web Scraping - Declan
Econometrics - Said
GitHub & Rmd - Rakeen

• Conclusion (If you don’t have the conclusion yet, please just present it on the presentation day.)
Null hypothesis: All regression coefficients equal 0. That is, the predictors cannot predict crises.
Alternative hypothesis: Not all regression coefficients equal 0. That is, the predictors can predict crises.


# Data collection and preparation

## Collect data

```{r}
african_crises <- read_csv("african_crises.csv", 
    col_types = cols(
        case = col_integer(), 
        cc3 = col_character(), 
        country = col_character(), 
        year = col_date(format = "%Y"), 
        systemic_crisis = col_integer(), 
        exch_usd = col_double(),
        domestic_debt_in_default = col_integer(), 
        sovereign_external_debt_default = col_integer(), 
        gdp_weighted_default = col_double(), 
        inflation_annual_cpi = col_double(),
        independence = col_integer(), 
        currency_crises = col_integer(), 
        inflation_crises = col_integer(), 
        banking_crisis = col_character()))
View(african_crises)
```

### Optional: Scrape data off web HTML


## Preliminary checks on data quality

## Diagnostics for relationshpis and strong interactions

## Apply remedial measures, if needed


# Reduction of number of explanatory variables (for exploratory observational studies)

## Determine several potentially useful subsets of explanatory variables; include known essential variables


# Model refinement and selection

## Investigate curvature and interaction effects more fully

# Model validation - 2-3 data sets